---
title: "Metabolomics"
output: html_document
date: "2025-11-02"
---
```{r}
library(dplyr)
```

### "P01" follow up visit circa ~2015 on ~ 500 adolescents (includes 223 of the 250 seen in 2012)

#Reading in data
```{r}
data_ses<-read.csv("/Users/jackielindstrom/Documents/ELEMENT/P01_SES_201910.csv") # 
data_met<-read.csv("/Users/jackielindstrom/Documents/ELEMENT/P01_Metab_Named_Logscale_202002.csv") #2015 P01 on ~ 500 adolescents (includes 223 of the 250 seen in 2012)
data_sex_age<-read.csv("/Users/jackielindstrom/Documents/ELEMENT/P01_Sex_and_Age_202202.csv")
data_maturation<-read.csv("/Users/jackielindstrom/Documents/ELEMENT/P01_Sexual_Maturation_201910.csv")
data_metP20<-read.csv("/Users/jackielindstrom/Documents/ELEMENT/P20_Metabolites_QCadj_KNN5_20171204.csv") # 2011: Untargeted metabolomics from fasting blood & targeted on amino acids and acylcarnitines ... not used for this project!
dim(data_met)
```

# Check: Sex=2: female, Sex=1: male?
```{r}
head(data_sex_age)
```

# Pubertal onset not computed in data_maturation data frame yet. Code to calculated it below:
```{r}
#initiate binary puberty onset column
data_maturation$pub_onset<-NA
#pub_onset=0 if both P2_1 and P2_2 for female =1
data_maturation$pub_onset<-ifelse(data_maturation$p2_1==1 &  data_maturation$p2_2==1 & data_maturation$sexo==2, 0, data_maturation$pub_onset)

#pub_onset=0 if both P2_1 and P2_3 for male =1
data_maturation$pub_onset<-ifelse(data_maturation$p2_1==1 &  data_maturation$p2_3==1 & data_maturation$sexo==1, 0, data_maturation$pub_onset)

#pub_onset=1 if either P2_1 or P2 _2 for female is greater than 1
data_maturation$pub_onset<-ifelse((data_maturation$p2_1>1 |  data_maturation$p2_2>1) & data_maturation$sexo==2, 1, data_maturation$pub_onset)

#pub_onset=1 if either P2 _1 and P2_ 3 for male is greater than 1
data_maturation$pub_onset<-ifelse((data_maturation$p2_1>1 |  data_maturation$p2_3>1) & data_maturation$sexo==1, 1, data_maturation$pub_onset)

table(data_maturation$pub_onset)
```


# Make new data frames for each with just etapa = T1 visits
```{r}
data_ses_T1 <- data_ses[data_ses$etapa == "T1", ]
data_sex_age_T1 <- data_sex_age[data_sex_age$etapa == "T1", ]
data_mat_T1 <- data_maturation[data_maturation$etapa == "T1", ]
```

# Recode 13x6 SES into numerical, with A/B= 6 being the best SES
```{r}
data_ses_T1$amai_13x6num <- dplyr::recode(data_ses_T1$amai_13x6,
                      "A/B" = 6,
                      "C+" = 5,
                      "C" = 4,
                      "D+" = 3,
                      "D" = 2,
                      "E" = 1)
```

# Recode 8x7 SES into numerical, with A/B= 7 being the best SES
```{r}

data_ses_T1$amai_8x7num <- dplyr::recode(data_ses_T1$amai_8x7,
                      "A/B" = 7,
                      "C+"= 6,
                      "C" = 5,
                      "C-" = 4,
                      "D+" = 3,
                      "D" = 2,
                      "E" = 1)
```

```{r}
hist(data_ses_T1$amai_8x7num)
```

```{r}
hist(data_ses_T1$amai_13x6num)
```

# Merge variables of interest into one data frame (use "FOLIO"). Save to CSV "merged_data.csv"
```{r}
# left join keeps all rows in data_met, but will put NA if nothing in age/ses/sex
merged_data <- data_met %>%
  left_join(select(data_ses_T1, foliocc, amai_13x6num, amai_8x7num), c("FOLIOCC" = "foliocc")) %>%
  left_join(select(data_sex_age_T1, FOLIOCC, Age_P01, AC_sexH), by = "FOLIOCC") %>%
  left_join(select(data_mat_T1, foliocc, pub_onset), by = c("FOLIOCC" = "foliocc"))

write.csv(merged_data, "merged_data.csv", row.names = FALSE)
head(merged_data)
```


# Trying one linear regression with one metabolite
```{r}
fit1 <- lm(Keto_14_0_1 ~ amai_8x7num + Age_P01 + pub_onset + AC_sexH, data = merged_data)
summary(fit1)
```

```{r}
# Select just metabolite columns from merged data
metabolite_names <- setdiff(names(merged_data), c("FOLIOCC", "amai_13x6num", "amai_8x7num", "Age_P01", "pub_onset", "AC_sexH"))

# Linear regression for each metabolite
results_list <- lapply(metabolite_names, function(met){
  formula <- as.formula(paste(met, "~ amai_8x7num + Age_P01 + pub_onset + AC_sexH"))
  fit <- lm(formula, data = merged_data)
  summary(fit)$coefficients  # get the coeff table
})

# Name list with metabolite names
names(results_list) <- metabolite_names

# Ex: see results for metabolite 55
results_list[[55]]
metabolite_names[[5]]

# count how many have sig p values for SES
pvals_amai <- sapply(results_list, function(res){
  res["amai_8x7num", "Pr(>|t|)"]
})
betas_amai <- sapply(results_list, function(res){
  res["amai_8x7num", "Estimate"]
})
```
```{r}

```

# Correct for FDR using benjamin- h
```{r}
FDR_amai <- p.adjust(pvals_amai, method = "BH")

all_results <- data.frame(
  metabolite = metabolite_names,
  beta =betas_amai,
  p.value = pvals_amai,
  FDR = FDR_amai
)
```


```{r}
sig_before_FDR <- subset(all_results, p.value < 0.05)
sig_before_FDR <- sig_before_FDR[order(sig_before_FDR$p.value), ]

print(sig_before_FDR)
write.csv(sig_before_FDR, "sig_before_FDR", row.names=FALSE)
```

```{r}
# FDR < 0.1
sig_FDR_10 <- subset(all_results, FDR < 0.1)
sig_FDR_10 <- sig_FDR[order(sig_FDR$FDR), ]

print(sig_FDR_10)
write.csv(sig_FDR_10, "sig_FDR_10.csv", row.names = FALSE)

# FDR < 0.2
sig_FDR_20 <- subset(all_results, FDR < 0.20)
sig_FDR_20 <- sig_FDR_20[order(sig_FDR_20$FDR), ]
print(sig_FDR_20)
```







